Last glacial cycle glacier erosion potential in the Alps

### 2020-12-30 articles/erosion/alps-erosion.tex
Draft Alps erosion intro, refs missing.

    The glacial landscape of the Alps has fascinated generations of explorers,
    artists, mountaineers and scientists with its diversity, including
    erosional features of all scales from high-mountain cirques, to steep
    glacial valleys and large overdeepenings. Using previous glacier modelling
    results, and modern observations of bedrock erosion under glaciers, we
    infer a distribution of potential glacier erosion in the Alps over the last
    glacial cycle from 120\,000 years ago to the present.
    %
    Depsite large uncertainties related to the climate history of the Alps and
    glacier erosion processes, the resulting modelled patterns of glacier
    erosion shows persistent features (hopefully). The cumulative imprint of
    the last glacial cycle shows peak erosion at the mouth of the large Alpine
    valleys where glaciers are modelled to have flown with the highest
    velocity. The modelled erosion pattern varies significantly through the
    glacial cycle, but surprisingly the total rate of glacier erosion is
    modelled to be relatively stable during the entire glacial cycle.  While
    glacial maxima lead to high erosion rates at low elevations, the
    high-elevation areas are typically preserved under cold-based ice during
    these periods, but are modelled to have experience mode intense erosion
    during periods of intermediate glaciation extent.
    %
    This result indicate that different landscapes of the same mountain range
    most likely correspond to different tmie periods, and explains the
    diversity of glacial landscapes in the Alps.

@@

    The glacial erosion landscape of the Alps has fascinated generations of
    explorers, artists, and scientists for centuries. In English, a non-Alpine
    language, the word “alpine” is now used to describe any glacially modified
    mountain landscape worldwide.

    Some regions are characterised by cirque glaciation, glacial valleys,
    or overland scouring. Other regions, including the Alps, present a variety
    of glacial erosional landforms. It is often assumed that the degree of
    glacial modification is a proxy for the duration of past glacier cover.

    While it has long been understood that glaciers leave a specific imprint on
    the landscape they lie on, observing and quantifying the long-term erosion
    and sedimentation processes has been a challenge. Glacier sliding
    velocities vary by several orders of magnitude, and erosion laws can be
    non-linear. Besides, cold-based glaciers have been shown to preserve
    landscape, but it remains unclear how sharp the erosional transition to
    cold-base ice is, and why some glaciated regions show traces of cold-base
    glaciations and others not.

    Here we apply a non-linear erosion law to previously published model
    results and analyse the patterns of modelled erosion rate and cumulative
    last glacial cycle erosion potential.

    Fig.~\ref{fig:landscape} -- Alpine glacial erosion landscape photos.

>>>>

    The glacial landscape of the Alps has fascinated generations of explorers,
    artists, mountaineers and scientists with its diversity, including
    erosional features of all scales from high-mountain cirques, to steep
    glacial valleys and large overdeepenings. Using previous glacier modelling
    results, and modern observations of bedrock erosion under glaciers, we
    infer a distribution of potential glacier erosion in the Alps over the last
    glacial cycle from 120\,000 years ago to the present.
    %
    Depsite large uncertainties related to the climate history of the Alps and
    glacier erosion processes, the resulting modelled patterns of glacier
    erosion shows persistent features (hopefully). The cumulative imprint of
    the last glacial cycle shows peak erosion at the mouth of the large Alpine
    valleys where glaciers are modelled to have flown with the highest
    velocity. The modelled erosion pattern varies significantly through the
    glacial cycle, but surprisingly the total rate of glacier erosion is
    modelled to be relatively stable during the entire glacial cycle.  While
    glacial maxima lead to high erosion rates at low elevations, the
    high-elevation areas are typically preserved under cold-based ice during
    these periods, but are modelled to have experience mode intense erosion
    during periods of intermediate glaciation extent.
    %
    This result indicate that different landscapes of the same mountain range
    most likely correspond to different time periods, and explains the
    diversity of glacial landscapes in the Alps.

@@

    The glacial erosion landscape of the Alps has fascinated generations of
    explorers, artists, and scientists for centuries. Indeed, its cultural
    impact is so far-reaching that in English, a non-Alpine language, the
    adjective ``alpine'' with non-capital ``a'' is commonly used to describe
    an Alpine-like, glacially modified, mountain landscape outside the Alps,
    while the proper noun ``Alps'' has been reused to nick-name glacier eroded
    mountain ranges of New Zealand, Japan, and others.

    Some mountain ranges are predominantly characterised by cirque glaciation
    (e.g. Uinta Mountains), glacial valleys (e.g. Putorana Plateau), or
    large-scale overdeepenings (e.g. Patagonia). But other regions, including
    the Alps, present a higher variety glacial erosional landforms, thereby
    bearing a different history of glacial erosion (Fig.~\ref{fig:landscape}).
    [The degree of glacial modification of a landscape has sometimes been used
    as a proxy for the duration of past glacier cover \aref, including for
    instance in geomorphological mapping \aref.] This paper challenges this
    assumption.

    While it has long been understood that glaciers leave a specific erosional
    imprint on the landscape, observing and quantifying the long-term erosion
    and sedimentation processes has been a challenge, not least due to the
    inaccessibility of glacier beds and the slow and stochastic nature of
    glacial erosion processes. More recent studies have moved away from process
    understanding in attempts to determine empirical erosion laws for an entire
    glacier or set of glaciers. \citet{Herman.etal.2015} collected sediment
    samples for \todo{X} months at the snout of Franz-Josef Glacier and mapped
    their chemical composition to different geological zones, and hence
    different glacier velocity, to construct a relationship between the
    reconstructed glacier basal velocity and the measured erosion rate.
    \citet{Koppes.etal.2015} \todo{check what they did}. \citet{Cook.etal.2020}
    \todo{check what they did}.

    These studies show that, while glacier and ice sheet velocities vary by
    several orders of magnitude, their relationship to erosion can be
    non-linear, yielding an even larger spectrum of variation for glacier
    erosion rates. Besides, cold-based glaciers have been shown to preserve
    landscape, but it remains unclear how sharp the erosional transition to
    cold-base ice is, and why some glaciated regions show traces of cold-base
    glaciations and others not.

    Here, a non-linear erosion law is applied to previously published model
    results and the patterns of modelled erosion rate and cumulative
    last glacial cycle erosion potential are analysed. The resulting
    uncertainties are expectantly enormous, yet the results are hoped to
    provide insights on the diversity of the Alpine glacial erosion landscape.

    \note{OK, maybe I need to do a literature review...}
